
% ============================================================
\section{Discussion}
\label{sec:discussion}
% ============================================================

\subsection{Summary of the Research Journey}

The path from classical BM theory to Margin Rank was non-linear. Table~\ref{tab:journey}
summarises the key conceptual transitions.

\begin{table}[H]
  \centering
  \begin{adjustbox}{width=\columnwidth}
  \begin{tabular}{lll}
    \toprule
    \textbf{Stage} & \textbf{Initial Assumption} & \textbf{What We Discovered} \\
    \midrule
    Classical BM    & Pataki bound gives $p^*$         & Bound inapplicable to asymmetric Softmax \\
    Warp Drive      & Softmax is just a nonlinearity   & Softmax absorbs $n$ constraints via null space \\
    Implicit Reg    & Nuclear norm oracle predicts $p^*$ & Norm-dimension paradox; lazy optimizer \\
    Margin Rank     & Need SDP relaxation              & Hard Bottleneck sweep is the correct oracle \\
    GPT-2           & All heads use similar capacity   & Heterogeneous landscape; 80--97\% redundancy \\
    \bottomrule
  \end{tabular}
  \end{adjustbox}
  \caption{Research journey: conceptual transitions and their causes.}
  \label{tab:journey}
\end{table}

\subsection{Limitations and Caveats}

\begin{itemize}
  \item \textbf{Static attention.} The Hard Bottleneck Predictor requires a fixed,
        pre-computed attention matrix. It cannot be applied during training without
        significant overhead.
  \item \textbf{Margin sensitivity.} The choice of $\gamma$ affects $p^*$; the theory does
        not yet provide a principled method for selecting $\gamma$ from first principles.
  \item \textbf{GPT-2 scale.} Our empirical profiling used GPT-2 (117M parameters). Whether
        the heterogeneous capacity landscape and 80\% redundancy figures hold for frontier
        models (GPT-4, LLaMA 3) remains an open empirical question.
  \item \textbf{Conjecture~\ref{conj:strict_saddle}.} The strict saddle property of the
        Softmax BM landscape at the corrected rank bound remains unproved.
\end{itemize}

% ============================================================
\section{Future Work}
\label{sec:future}
% ============================================================

Several open problems emerge directly from this work:

\begin{enumerate}
  \item \textbf{Completing the strict saddle proof.} The local surjectivity argument for the
        Softmax Jacobian restricted to the active constraint manifold is the key missing piece
        for a full proof of Conjecture~\ref{conj:strict_saddle}.
  \item \textbf{Non-commutative gradient flow.} Proving or disproving the nuclear norm
        implicit regularization conjecture in the non-commutative Softmax BM case.
  \item \textbf{Frontier-model profiling.} Applying the Hard Bottleneck Predictor to LLaMA-3
        and measuring whether Margin Rank scales sub-linearly with model depth.
  \item \textbf{Training-time integration.} Designing a differentiable Margin Rank proxy
        that can regularize $d_{\text{head}}$ during training, enabling Heterogeneous
        Transformer training from scratch.
  \item \textbf{Precision-dependent formula fitting.} A formula
        \begin{equation}
        \begin{aligned}
            p^*(\varepsilon, \text{eff}_r, \tau, L)
        \end{aligned}
        \end{equation}
        incorporating precision, effective rank, temperature, and
        depth, fitted using \texttt{scipy.optimize.curve\_fit} with confidence intervals.
\end{enumerate}

% ============================================================
\section{Conclusion}
\label{sec:conclusion}
% ============================================================

This report has traced a complete research arc from classical Burer-Monteiro theory to a new
framework, Margin Rank Theory, that mathematically characterises the true structural
capacity of Transformer attention heads.

Our central results are:
\begin{enumerate}
  \item The Barvinok-Pataki bound, while empirically verified in the symmetric SDP setting,
        is fundamentally inapplicable to Softmax attention due to the asymmetric geometry and
        the null space of the Softmax Jacobian.
  \item The Softmax operator converts $m$ linear equality constraints into margin
        inequalities, absorbing $n$ row-sum constraints and enabling an order-of-magnitude
        dimensional compression (the Geometric Warp Drive: $p^*=2$ for an Identity matrix of
        any size).
  \item The Margin Rank of an attention head, the minimum embedding dimension required to
        geometrically separate targets from distractors, is the correct measure of head
        capacity, governed by topological complexity, not linear rank.
  \item Real GPT-2 attention heads operate at 80--97\% dimensional redundancy, with clear
        heterogeneous capacity requirements across heads in the same layer.
\end{enumerate}

These findings provide rigorous mathematical justification for the development of
\textbf{Heterogeneous Transformers}: future architectures where head dimensions are
dynamically allocated by Margin Rank profiling, promising substantial reductions in
computational footprint without sacrificing semantic expressivity.

\section*{Acknowledgements}

This research was conducted as part of the IE210 Optimization Modelling course at IIT
Bombay, 2026. The empirical experiments were implemented in PyTorch and CVXPY, with MOSEK
as the convex solver baseline. GPT-2 attention tensors were extracted using the Hugging Face
Transformers library.

% ============================================================
% \end{document}